A Switching Planner for Combined Task and Observation Planning

### 2010-11-25 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/switching.tex
switching is starting to mature

We describe our continual planning approach that switches between
sequential and contingent sessions. Continual planning and execution
proceeds in a deterministic-action POMDP described in DTPDDL. During a
sequential session, planning is performed by a ``classical''
system,\footnote{That is, a system designed to solve fully observable
deterministic tasks.}  and execution proceeds according to the {\em
trace} computed by that system. Taking the form of a classical plan,
the {\em trace} specifies a sequence of POMDP actions that achieve the
agent's objectives in a deterministic approximation, i.e., {\em
determinisation}, of the problem at hand. More precisely, the trace is
an interleaved sequence of POMDP actions and {\em assumptive}
actions. The latter correspond to assumptions the planner makes about
the truth value of {\em runtime state variables} -- e.g. that a box of
cornflakes is located in the kitchen at the third plan step. Another
kind of assumption, called an {\em applicability} assumption, is made
when the trace includes an action $\action$ that is not applicable
with probability $1$ at the belief-state
\bstate\ that the system is projected to be in when \action\ is executed, 
i.e., $\exists\state\in\states
\bstate(\state) > 0$ and $\poss(\action)\not\subseteq\state$. By
scheduling
\action,  the serial planner makes an assumption about the
observability of the precondition $\poss(\action)$.

@@

Our approach uses sequential planning to compute a linear plan, and
one trace of that plan's execution. The latter encapsulates
assumptions about: (1) the true underlying state, (2) how the plan's
execution will progress, and (3) the possibility of the agent
eventually holding strong beliefs about the truth values of specific
state propositions.

In a deterministic-action POMDP, all the uncertainty in state and
action is expressed in the $(\pp{:init})$ declaration. Our approach
uses the structure of that, as it occurs in the problem description,
to define a set of state-assumptions available to sequential planning.
%%
Writing \#\ if the value of a proposition is unspecified in an
assumption, for our $\pp{DORA}$ example we have the following
assumptions:

@@

\noindent A valid assumption in our setting corresponds to a {\em
relaxed} visitation of the root term of $(\pp{:init})$. In this relaxed
case, a conjunctive term is visited iff its atomic subterms are
visited, and zero or more of its immediate probabilistic subterms are
visited. The starting state, $\state_0$, for sequential planning is
the unique abstract state with probability $1.0$ -- e.g., for
$\pp{DORA}$:

@@

In order to make assumptions available to the sequential planner, we
add one {\em assumptive} action $\assumptiveS{i}$ to the problem for
each element, $p_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. The physics of these actions is as
follows. $\assumptiveS{i}$ can be executed if no $\assumptiveS{j}$,
$j \neq i$, has been executed from the same probabilistic term, and,
either $(\pp{probabilistic}~..p_i~(T_i)..)$ is in the root conjunct,
or it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
Executing $\assumptiveS{i}$ in $s$ effects a transition to a successor
state $s^{T_i}$ with probability $p_i$, and $s^\bot$ with probability
$1 - p_i$. Here, $s^{T_i}$ is the union of $s$ with atomic terms from
$T_i$ and conflicting propositions from $s$ removed. For example, if
the action adds $(=(\pp{is-in}~\pp{box})~\pp{office})$, then it must
delete $(=(\pp{is-in}~\pp{box})~\#)$. State $s^\bot$ is a sink, where
$\forall\action\in\actions$ $\reward(\state^\bot, \action) = 0$ and
$\transProb(\state^\bot, \action, \state^\bot) = 1$.

Because $\state_0$ is abstract, and because actions can have
preconditions that are not uniformly true in the underlying
(resp. abstract) belief-state, we must make two further adjustments to
the problem posed to the sequential planner. First, if
$\action\in\actions$ has a precondition that is not true with
probability $1$ in $\bstate_0$, then before it is executed blindly,
the sequential continual planning switches to a decision-theoretic
session, discussed in detail below. Overall, by scheduling $\action$
the sequential planner makes a naive assumption about the possibility
of the agent eventually ``committing'' to the truth values of specific
state propositions in $\poss(\action)$. The role of decision-theoretic
plans is to examine the validity of assumptions about the runtime
state of sequential planning. Second, if in the current state \state\
the value of $\prop$ is $\#$ ---because a truth value is not specified
in $\state_0$ and no assumption is made about it in the plan
to \state--- then if $\prop \in \poss(\action)$, or if $\prop$ is the
subject of a positive or negative effect of $\action$, we have that
action $\action$ is not applicable in \state.

Finally, we come to valuing sequential plans, and therefore giving a
criteria for the sequential planner to optimise. Where $p_i$ is the
probability that the $i^{th}$ sequenced action does not transition to
$s^\bot$, we value a trace as follows:

\begin{equation}\label{eq:tracevalue}
V(s_0, a_0, s_1, a_1,.., s_N) =  \prod_{i=1..N-1} p_i \sum_{i=1..N-1} \reward(\state_i, \action_i)
\end{equation}

\noindent The optimal trace given a plan, i.e., the non-assumptive
actions in sequence, therefore has value:

\[
V^* = \max_{p_1, .., p_n} \prod_{i=1..N-1} p_i \sum_{i=1..N-1}
\reward(\state_i, \action_i),
\]

\noindent which is equal to the maximal contribution a trace of the plan can
make to Eq~\ref{eq:expectedvalue}. %% In other words, writing $V_i$ for
%% the value of the $i^{th}$ trace of the plan expressed in $s_0, a_0,
%% s_1, a_1,.., s_N$, we have
%%
%% \[
%% V^* = \arg\min_{V_i} \Expect \bigg{[} 
%% \sum_{t=0}^{N-1}  \reward(\bstate_t, \pp{PLAN}_t) \mid \pp{PLAN}, \bstate_0
%% \bigg{]} - V_i.
%% \]
%%
In goal directed problems, where the only non-zero reward is received
at the first transition to a goal state, the Eq~\ref{eq:tracevalue}
criteria corresponds exactly to that of
\system{FFR$_a$}~\cite{yoon:etal:2007}. From the perspective of
gradient-based reinforcement learning systems, such as William's
REINFORCE and GPOMDP variants for planning~\cite{olivier:doug:2009},
under reasonable assumptions the optimal trace identifies a maximal
gradient step from a uniformly random soft-max policy. Finally, it is
worth clarifying that although the above discussion has been in terms
of a trace of length $N$, there is nothing in our approach that
artificially limits the length of candidate sequential plans.

>>>>

We describe our continual planning approach that switches between
sequential and contingent sessions. Continual planning and execution
proceeds in a deterministic-action POMDP described in DTPDDL. During a
sequential session, planning is performed by a ``classical''
system,\footnote{That is, a system designed to solve fully observable
deterministic tasks.}  and execution proceeds according to the {\em
trace} computed by that system. Taking the form of a classical plan,
the {\em trace} specifies a sequence of POMDP actions that achieve the
agent's objectives in a deterministic approximation, i.e., {\em
determinisation}, of the problem at hand. More precisely, the trace is
an interleaved sequence of POMDP actions and {\em assumptive}
actions. The latter correspond to assumptions the planner makes about
the truth value of {\em runtime state variables} -- e.g. that a box of
cornflakes is located in the kitchen at the third plan step. Another
kind of assumption, called an {\em applicability} assumption, is made
when the trace includes an action $\action$ that is not applicable
with probability $1$ at the belief-state
\bstate\ that the system is projected to be in when \action\ is executed, 
i.e., $\exists\state\in\states\;
\bstate(\state) > 0$ and $\poss(\action)\not\subseteq\state$. By
scheduling
\action,  the serial planner makes an assumption about the
observability of the precondition $\poss(\action)$.

@@

A sequential session uses a classical planner to compute a trace. The
latter encapsulates assumptions about: (1) the true underlying state,
(2) how the plan's execution will progress, and (3) the possibility of
the agent eventually holding strong beliefs about the truth values of
specific state propositions. Here we describe the deterministic
planning problem, derived from the DTPDDL model, that admits plans
which correspond to traces.

In a deterministic-action POMDP all the uncertainty in state and
action is expressed in the $(\pp{:init})$ DTPDDL declaration. Our
approach uses the structure of that, as it occurs in the problem
description, to define a set of state-assumptions available to
sequential planning.
%%
Writing \#\ if the value of a proposition is unspecified, for
$\pp{DORA}$ we have the following assumptions:

@@

\noindent In detail, a valid assumption in our setting corresponds to a {\em
relaxed} visitation of the root term of $(\pp{:init})$. In this
relaxed case, a conjunctive term is visited iff its atomic subterms
are visited, and zero or more of its immediate probabilistic subterms
are visited. The starting state, $\state_0$, for sequential planning
is an {\em abstract} state that captures the unique assumption which
has probability $1.0$. In $\pp{DORA}$, that is:

@@

In order to make assumptions available to the sequential planner, we
add one {\em assumptive} action $\assumptiveS{i}$ to the problem for
each element, $p_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. The physics of these actions is as
follows. $\assumptiveS{i}$ can be executed if no $\assumptiveS{j}$,
$j \neq i$, has been executed from the same probabilistic term, and,
either $(\pp{probabilistic}~..p_i~(T_i)..)$ is in the root conjunct,
or it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
Executing $\assumptiveS{i}$ in $s$ effects a transition to a successor
state $s^{T_i}$ with probability $p_i$, and $s^\bot$ with probability
$1 - p_i$. Here, $s^{T_i}$ is the union of $s$ with atomic terms from
$T_i$ and conflicting propositions from $s$ deleted. For example, if
the action adds $(=(\pp{is-in}~\pp{box})~\pp{office})$, then it must
delete $(=(\pp{is-in}~\pp{box})~\#)$. State $s^\bot$ is an added sink,
where $\forall\action\in\actions$ $\reward(\state^\bot, \action) = 0$
and $\transProb(\state^\bot, \action, \state^\bot) = 1$.

For sequential planning the operators from the POMDP model are
available with semantics that accommodate the above abstraction as
follows. In the deterministic model a proposition $\prop$ can be
thought to have a ternary interpretation at a state \state, as either
{\em true}, written $\prop\in\state$, {\em false}, written
$\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
notation written $\prop=\#\in\state$. For example, in $\pp{DORA}$
$(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
$\action\in\actions$ in the deterministic model, if
$\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
or negative effect of $\action$, then $\action$ is not applicable
in \state\ if $\prop=\#\in\state$. Indeed, first an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment could be modified by POMDP operators. Lastly, we
address the switching semantics of action execution. If
$\action\in\actions$ has a precondition that is not true with
probability $1$ in the underlying belief $\bstate$, then before it is
executed blindly, the system switches to a contingent session,
discussed in detail in the following section.

We now describe the optimisation criteria given to the classical
planner. Where $\prob_i$ is the probability that the $i^{th}$
sequenced action $\action_i$ does not transition to $\state^\bot$, we
define the value of a trace
$\state_0, \action_0, \state_1, \action_1,.., \state_N$ from the
deterministic model to equal:

\begin{equation}\label{eq:tracevalue}
V(\state_0, \action_0, \state_1, \action_1,.., \state_N) =  \prod_{i=1..N-1} \prob_i \sum_{i=1..N-1} \reward(\state_i, \action_i)
\end{equation}

\noindent The optimal trace given a plan, i.e., the sequence of non-assumptive
actions from the trace, therefore has value:

\[
V^* = \max_{\prob_1, .., \prob_n} \prod_{i=1..N-1} \prob_i \sum_{i=1..N-1}
\reward(\state_i, \action_i),
\]

\noindent which is equal to the maximal additive contribution a trace of the plan can
make to Eq~\ref{eq:expectedvalue}, taking an FSC that corresponds to
the sequential plan.  In goal directed problems, where the only
non-zero reward is received at the first transition to a goal state,
the Eq~\ref{eq:tracevalue} criteria gives us the behaviour
of \system{FFR$_a$}~\cite{yoon:etal:2007}. From the perspective of
gradient-based reinforcement learning systems, such as William's
REINFORCE and GPOMDP variants for planning~\cite{olivier:doug:2009},
under reasonable assumptions the optimal trace identifies a maximal
gradient step from a uniformly random soft-max policy. Finally, it is
worth clarifying that although the above discussion has been in terms
of a trace of length $N$, there is nothing in our approach that
artificially limits the length of candidate sequential plans.
